---
title: "ENVS 193DS Final"
author: "Michael Torbensen"
date: June 05, 2026
format:
  html:
    toc: TRUE
    execute: 
      message: FALSE
      warning: FALSE
---

[Github Repository](https://github.com/michaeltorbensen/ENVS-193DS_spring-2026_final.git)

# Set-Up

```{r}

#| label: loading in packages and data
library(tidyverse)
library(janitor)
library(here)
library(readxl)
library(ggeffects)
library(MuMIn) # model selection
library(DHARMa)
library(gtsummary)
library(rstatix)

# reading in data 
bike_data <- read.csv(here::here("data/personal_data.csv"))
nest_data <- read.csv(here::here("data/nest_data_final.csv"))

```

# Problem 1. Research Writing 

## a. Transparent statistical methods

In part 1, my co-workers used a generalized linear model (GLM) to show if distance to water edge predicts the probability of American Avocet microhabitat use. In part 2, my co-workers used a chi-squared test to see if there was a relationship between bird species and habitat types. 

## b. Suggestions for rewriting

#### Part 1 

The distance between the habitat and water line was a significant predictor for the American Avocet using the microhabitat or not. 

#### Part 2

The distribution of bird species varied significantly across habitat types (managed pond, non-tidal marsh, salt pond, tidal marsh, and tidal mudflat). Results suggest that the studied bird species (American Avocet, Forster’s Tern, Black-necked Stilt) prefer one habitat over another or may have different habitat uses. 


## c. Further analysis

Another variable that can influence the probability of American Avocet microhabitat use is native plant cover (continuous variable) in the area of study. It can be measured as percent cover 0–100% by visual estimation within fixed plots. This factor would influence the habitat use as higher native plant cover may reduce the accessible foraging habitat and visibility Americam Avocets need. 

# Problem 2. Data Analysis

## a. Clean and wrangle 

```{r}
#| label: cleaning-and-wrangling-data

# Creating new object 
nests_clean <- nest_data |> 
  # Creating new columns to store taxonomic names 
  mutate(ant_species = case_match(
    ant_species,
    "AB" ~ "Crematogaster sjostedti",
    "RRB" ~ "Crematogaster mimosae",
    "TP" ~ "Tetraponera penzigi",
    "BBR" ~ "Crematogaster nigriceps",
  )) |>
  # Filtering to include certain species (C. mimosae, C. sjostedti, and C. nigriceps)
  filter(ant_species %in% c("Crematogaster mimosae", 
                            "Crematogaster sjostedti", 
                            "Crematogaster nigriceps")) |> 
  # Changing ant species order 
  mutate(ant_species = as_factor(ant_species),
         ant_species = fct_relevel(ant_species,
                                       "Crematogaster sjostedti",
                                       "Crematogaster mimosae",
                                       "Crematogaster nigriceps")) |> 
  # Selecting columns of interest for research questions
  select(height_cm, ant_species, case_control)

# Display new data object 
str(nests_clean)

# Display 10 random rows
slice_sample(nests_clean, n = 10)

```

# b. Response variable 

In a biological context, the 0's and 1's represent a binary outcome of yes (1) or no (0). In the context of this study, a 1 represents the presence of a nest in a tree and 0 represents the absence of nests on a tree. 

# c. Purpose of study

Tree height (numerical continuous variable), measured in cm, is one of the main predictors. Determined by the results section in paragraph 2, as tree height increase, the probability of nest occurrence would increase because height can protect the nests from ground predators. Yet, this is also dependent on what ant species are present in the tree, as stated in paragraph 3 of the results section. 

As established in paragraph 3 of the results section, the other main predictor is the species of ant, a categorical variable including *C. mimosae*, *C. sjostedti*, and *C. nigriceps*, present in the tree. The following information comes from paragraphs 3 and 4 of the introduction section. Since ants normally prey on nests in acacia trees, it's usual that nest quantity will be higher in tree with less aggressive ant species like *C. sjostedti*. Nevertheless, since the acacia ants protect the tree fro, varying nest predators, it's predicted that the quantity of nests will be higher in trees with more aggressive symbionts like *C. mimosea* and *C. nigriceps*. 

## d. Table of models 

| Model number | Ant Species |   Tree Height   | Interaction?                |
|--------------|:-----------:|:---------------:|:---------------------------:| 
| 1            |      X      |        X        |      no interaction (+)     | 
| 2            |      X      |        X        |      interaction (*)        |

## e. Run the models 

```{r}

#| label: running models 

# running model 1 (no interaction)
nest_mod1 <- glm(case_control ~ ant_species + height_cm, # "+" = no interaction
                 data = nests_clean, # from cleaned data set
                 family = "binomial") # binomial data set

# running model 2 (interaction)
nest_mod2 <- glm(case_control ~ ant_species * height_cm, # "*" = interaction
                data = nests_clean, # from cleaned data set
                family = "binomial") # binomial data set

```

