---
title: "ENVS 193DS Final"
author: "Michael Torbensen"
date: June 05, 2026
format:
  html:
    toc: TRUE
    execute: 
      message: FALSE
      warning: FALSE
---

[Github Repository](https://github.com/michaeltorbensen/ENVS-193DS_spring-2026_final.git)

# Set-Up

```{r}
#| label: loading in packages and data

library(tidyverse)
library(janitor)
library(here)
library(readxl)
library(ggeffects)
library(MuMIn) # model selection
library(DHARMa)
library(gtsummary)
library(rstatix)

# reading in data 
bike_data <- read.csv(here::here("data/personal_data.csv"))
nest_data <- read.csv(here::here("data/nest_data_final.csv"))

```

# Problem 1. Research Writing 

## a. Transparent statistical methods

In part 1, my co-workers used a generalized linear model (GLM) to show if distance to water edge predicts the probability of American Avocet microhabitat use. In part 2, my co-workers used a chi-squared test to see if there was a relationship between bird species and habitat types. 

## b. Suggestions for rewriting

### Part 1 

The distance between the habitat and water line was a significant predictor for the American Avocet using the microhabitat or not. 

### Part 2

The distribution of bird species varied significantly across habitat types (managed pond, non-tidal marsh, salt pond, tidal marsh, and tidal mudflat). Results suggest that the studied bird species (American Avocet, Forster’s Tern, Black-necked Stilt) prefer one habitat over another or may have different habitat uses. 


## c. Further analysis

Another variable that can influence the probability of American Avocet microhabitat use is native plant cover (continuous variable) in the area of study. It can be measured as percent cover 0–100% by visual estimation within fixed plots. This factor would influence the habitat use as higher native plant cover may reduce the accessible foraging habitat and visibility Americam Avocets need. 

# Problem 2. Data Analysis

## a. Clean and wrangle 

```{r}
#| label: cleaning data 

# Creating new object 
nests_clean <- nest_data |> 
  # Creating new columns to store taxonomic names 
  mutate(ant_species = case_match(
    ant_species,
    "AB" ~ "Crematogaster sjostedti",
    "RRB" ~ "Crematogaster mimosae",
    "TP" ~ "Tetraponera penzigi",
    "BBR" ~ "Crematogaster nigriceps",
  )) |>
  # Filtering to include certain species (C. mimosae, C. sjostedti, and C. nigriceps)
  filter(ant_species %in% c("Crematogaster mimosae", 
                            "Crematogaster sjostedti", 
                            "Crematogaster nigriceps")) |> 
  # mutating data 
  mutate(ant_species = as_factor(ant_species), # making sure data is written out as text
         ant_species = fct_relevel(ant_species, # making sure species will be listed in this order
                                       "Crematogaster sjostedti",
                                       "Crematogaster mimosae",
                                       "Crematogaster nigriceps")) |> 
  # Selecting columns of interest for research questions
  select(height_cm, ant_species, case_control)

# Display new data object 
str(nests_clean)

# Display 10 random rows
slice_sample(nests_clean, n = 10)

```

# b. Response variable 

In a biological context, the 0's and 1's represent an outcome occurring (1) or an outcome not occurring (0). In the context of this study, a 1 represents the presence of a nest in a tree and 0 represents the absence of nests on a tree. 

# c. Purpose of study

Tree height (numerical continuous variable), measured in cm, is one of the main predictors. Determined by the results section in paragraph 2, as tree height increase, the probability of nest occurrence would increase because height can protect the nests from ground predators. Yet, this is also dependent on what ant species are present in the tree, as stated in paragraph 3 of the results section. 

As established in paragraph 3 of the results section, the other main predictor is the species of ant, a categorical variable including *C. mimosae*, *C. sjostedti*, and *C. nigriceps*, present in the tree. The following information comes from paragraphs 3 and 4 of the introduction section. Since ants normally prey on nests in acacia trees, it's usual that nest quantity will be higher in tree with less aggressive ant species like *C. sjostedti*. Nevertheless, since the acacia ants protect the tree fro, varying nest predators, it's predicted that the quantity of nests will be higher in trees with more aggressive symbionts like *C. mimosea* and *C. nigriceps*. 

## d. Table of models 

| Model number | Ant Species |   Tree Height   | Interaction?                |
|--------------|:-----------:|:---------------:|:---------------------------:| 
| 1            |      X      |        X        |      no interaction (+)     | 
| 2            |      X      |        X        |      interaction (*)        |

## e. Run the models 

```{r}
#| label: running models
 
# model 1 (no interaction)
nest_mod1 <- glm(case_control ~ ant_species + height_cm, # no interaction (+)
                 data = nests_clean, # data from cleaned nest data set
                 family = "binomial") # binomial data set

# model 2 (interaction)
nest_mod2 <- glm(case_control ~ ant_species * height_cm, # interaction (*)
                data = nests_clean, # data from cleaned nest data set
                family = "binomial") # binomial data set

```

## f. Select the best model

```{r}
#| label: Akaike’s Information Criterion

AICc(nest_mod1,
     nest_mod2) |>
  arrange(AICc)

```

As concluded by Akaike’s Information Criterion (AIC), the best model for predicting the probability of bird nest presence is the interaction model between tree height and ant species since it has the lowest AIC. 

## g. Check the diagnostics 

```{r}
#| label: model diagnostics 

# checking diagnostics using simulated residuals from DHARMa package 
plot(
  simulateResiduals(nest_mod2)) # using interaction model 2

```

## h. Visualize the model predictions 

### Generating model predictions 

```{r}
#| label: model predictions

# object for prediction
preds <- ggpredict(
  nest_mod2,
  # denoting which predictors to use 
  terms = c("height_cm [74:598 by = 1]", "ant_species")) |> 
  # renaming columns for figure 
  rename(height_cm = x, 
         ant_species = group)
```

### Creating figure

```{r}
#| label: generating figure using ggplot

# base layer: ggplot
ggplot(data = nests_clean, # data from cleaned nest data 
       # x-axis: tree height
       # y-axis: nest probability
       mapping = aes(x = height_cm,
                     y = case_control)) + 
  # first layer: individual data points 
  geom_point(mapping = aes(color = ant_species), # color by ant species
             shape = 20) + # shape = small circle 
  # second layer: trend line based on predictions
  geom_line(data = preds, 
            # x-axis: tree height
            # y-axis: probability
            # colored by species
            mapping = aes(x = height_cm,
                          y = predicted,
                          color = ant_species)) +
  # third layer: 95% confidence interval ribbon
  geom_ribbon(data = preds, # data from model predictions
              # x-axis: tree height
              # y-axis: probability
              # colored by species
              mapping = aes(x = height_cm,
                            y = predicted,
                            ymin = conf.low,
                            ymax = conf.high, 
                            fill = ant_species),
                        alpha = 0.4) + # changing transparency of ribbon
  # making one panel for each species
  facet_wrap(~ ant_species) + 
  # customizing colors to categorize by species
  scale_color_manual(values = c(
    "Crematogaster mimosae" = "#c91a34",
    "Crematogaster sjostedti" = "#fa8b14", 
    "Crematogaster nigriceps" = "#ffc421"))  +
  # customizing colors to categorize by species 
  scale_fill_manual(values = c(
    "Crematogaster mimosae" = "#c91a34",
    "Crematogaster sjostedti" = "#fa8b14", 
    "Crematogaster nigriceps" = "#ffc421")) + 
  # labeling x- and y-axis
  labs(x = "Tree Height (cm)", 
       y = "Probability of Bird Nesting In Tree") + 
  # changing theme from default 
  theme(panel.background = element_rect(fill = "transparent", # no panel background
                                      colour = "darkblue", # dark blue outline
                                      linewidth = 1), # width of outline
        panel.grid = element_blank(), # taking out grid lines
        axis.ticks = element_line(linewidth = 1), # major tick marks on x- and y-axis
        legend.position = "none") # removing legend from panel

```

## i. Write a caption for your figure.

** Figure 1. Probability of bird nest in an acacia tree changes due to varying tree height and ant species.** Plots vary by ant species (seen at top tabs) as well as color (**C. sjostedti** = orange, **C. mimosae** = red, **C. nigriceps** = yellow). The points represent individual nesting observations, the ribbons expanding from the line are 95% confidence intervals, and the trend line represents the models predictions of whether or not a nest is present given the tree height (cm) and ant species. The probability of a bird nest in an acacia tree is either 1 (the presence of a nest(s) in the tree) or 0 (absence of nest(s) in the tree). Data sourced from Lujan, Ema et al. 2023. “Symbiotic acacia ants drive nesting behavior by birds in an African savanna.” Biotropica 55:6. doi: 10.1111/btp.13276.

## j. Calculate model predictions 

```{r}
#| label: model prediction calculation

# ggpredict for model predictions at 600cm height only
ggpredict(nest_mod2, # data from prediction model 2
          terms = c("height_cm [600]",
                    "ant_species")) 

```

## k. Interpret results 

The predicted probability of nest occurrence at 600 cm tree height (tallest tree in the data set) for **C. sjostedti** is 0.2 (95% CI: [0.00, 0.99]), for **C. mimosae** is 0.74 (95% CI: [0.48, 0.90]), and for **C. nigriceps** was 1.00 (95% CI: [1.00,1.00]. Tree height has a positive relationship with probability of bird nesting in trees occupied by all three ant species, though the strength and shape of this relationship differed vastly. The relationship was most gradual for *C. mimosae*, most abrupt for *C. nigriceps* (which showed a near-threshold response at low heights), and weakest with greatest uncertainty for *C. sjostedti*. Biologically speaking, more aggressive ant symbionts like **C. nigriceps** and **C. mimosae** are known to defend their host tree from nest predators like snakes and raptors. Even beyond aggression, **C. nigriceps** physically remodels tree growth by pruning apical buds and producing denser canopies. Birds appear to exploit this defensive behavior and extra suitable habitat to reduce nest predation risk. 

# Problem 3. Affective and exploratory visualizations 

## a. Comparing visualizations 

### Differences 

My visualization in homework 2 did not show any trend that was recognizable since there were not that many data points, but the next visualization (affective visualization) showed a decent pattern in both trials and trips as there were many more observations to work with. The first scatter plot I made in homework 2 showed that my duration of trip #2 was faster than my trip #1 duration while the affective visualization showed the opposite. 

### Similarities 

I do see some similarities in between the two visualizations. One being the effect that wind had on my travel duration (stronger wind makes = longer duration) as well as the effect that stop lights had on my travel duration (more stop lights = longer duration). 

### Patterns


### Feedback


